---
title: "DOAS Tutorial: Analyzing Real Survey Data and Running Simulations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DOAS Tutorial: Analyzing Real Survey Data and Running Simulations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The **Double Observer Adjusted Survey (DOAS)** framework reduces the effort required to estimate wild animal abundance by combining standard block counts with a subset of Double Observer (DO) counts. 

This tutorial covers the complete workflow of the `DOAS` package. You will learn how to:
1. **Analyze real survey data** using the `doas()` function.
2. **Simulate animal populations and surveys** using `simulate_pop()` and `simulate_survey()`.
3. **Run power analyses** using `doas_power()` to plan your field sampling efforts.
4. **Understand the underlying base estimators**.

---

## 1. Analyzing Survey Data (`doas`)

Your data needs to be summarized at the **sector level**. The final `data.frame` passed to the `doas` function must contain exactly one row per survey sector.

### Required Columns:
1. **`sector_id`**: A unique identifier for the sector.
2. **`total_count`**: The raw total number of individuals counted in that sector.
3. **`detected_groups`**: The total number of unique groups detected in that sector.

### Double Observer (DO) Columns (Optional):
For sectors that received a Double Observer survey, provide the following columns (leave them as `NA` for standard Block sectors):
4. **`groups_both`**: How many groups were seen by **both** observers.
5. **`groups_obs1`**: How many groups were seen by **Observer 1 only**.
6. **`groups_obs2`**: How many groups were seen by **Observer 2 only**.

> **Note:** The `doas()` function automatically calculates the mean group size internally (`total_count / detected_groups`). You do not need to provide it.

### Example Data

Let's imagine you surveyed 6 valleys. You did a standard block count in 4 of them, but you deployed two independent observers in 2 of the valleys.

```{r setup-real-data}
library(DOAS)

my_survey_data <- data.frame(
  sector_id = c("Valley_A", "Valley_B", "Valley_C", "Valley_D", "Valley_E", "Valley_F"),

  # Total animals seen by anyone in the sector
  total_count = c(55, 30, 80, 100, 20, 0),

  # Total unique groups seen
  detected_groups = c(8, 5, 12, 10, 2, 0),

  # Mark-Recapture data for DO sectors
  groups_both = c(6, 3, NA, NA, NA, NA),
  groups_obs1 = c(1, 1, NA, NA, NA, NA),
  groups_obs2 = c(1, 1, NA, NA, NA, NA)
)

print(my_survey_data)
```

### Running the Estimator

You can run the estimation using different observer assumptions (`obs_type`):
- `"independent"`: Default. Assumes observers scanned independently (uses Chapman estimator).
- `"dependent"`: Assumes Observer 1 points out animals to Observer 2, and Observer 2 only records what Observer 1 missed (Zippin estimator).
- `"bayesian"`: Uses Monte Carlo simulation from Conjugate Beta-Binomial posteriors.

```{r run-doas}
# Default Hybrid estimation with Independent observers
results <- doas(my_survey_data, method = "hybrid", obs_type = "independent")

summary(results)
```

You can view the specific sector-by-sector estimates directly from the results object:

```{r view-sectors}
knitr::kable(results$sector_estimates, digits = 2)
```

---

## 2. Simulating Populations and Surveys

To test the method or plan study designs, the DOAS package includes functions to simulate realistic cluster-sampled populations.

### Simulating a Population (`simulate_pop`)

`simulate_pop()` distributes a simulated population into logical sectors and groups.

```{r sim_pop}
set.seed(42)
# Simulate ~1000 individuals, mean group size 10, spread across 15 sectors
simulated_population <- simulate_pop(
  pop_size = 1000,
  mean_group = 10,
  tot_areas = 15
)

# Total simulated individuals
simulated_population$tot_pop
```

### Simulating a DOAS Survey (`simulate_survey`)

`simulate_survey()` takes the "true" population and applies detection probabilities to mimic field data collection.

```{r sim_survey}
# Simulate field observers surveying the population
survey_data <- simulate_survey(
  sim_pop = simulated_population$sim_pop,
  det_p1 = 0.6, # Observer 1 spots ~60% of groups
  intra_group_det = 1.0, # Observers count individuals within groups perfectly (no bias)
  det_group = 0.5, # Group size scaling factor
  obs_effect = 0.75, # Observer 2 is 75% as effective as Observer 1
  detvar = 0.2, # Variance in detection across different sectors
  surv_areas = 5 # Out of 15 areas, 5 will get Double Observers
)

head(survey_data)
```

This output is automatically formatted and ready to be plugged into `doas()`!

```{r run-sim-doas}
# We use 'global' method to pool detection probabilities, which is
# recommended to avoid small-sample biases on sparse sectors.
sim_results <- doas(survey_data, method = "global", obs_type = "independent")
summary(sim_results)
```

### Advanced: The Bayesian Approach

When observer detections are extremely low, classic estimators can sometimes produce negative estimates or fail. You can pass `obs_type = "bayesian"` to use a rigorous Monte Carlo Beta-Binomial framework that respects natural boundaries (i.e. you cannot estimate fewer animals than you physically saw).

```{r run-sim-doas-bayes}
bayes_results <- doas(survey_data, method = "global", obs_type = "bayesian")
summary(bayes_results)
```

---

## 3. Power Analysis (`doas_power`)

If you are planning a field season, you can use `doas_power()` to understand how many DO surveys you need to achieve reliable confidence intervals and minimize bias.

`doas_power()` runs multiple simulated surveys under the hood and reports metrics like Relative Bias, Root Mean Squared Error (RMSE), and 95% CI Coverage.

```{r doas_power}
# Run a quick power analysis testing 2 vs 5 DO sectors
# Note: For real power analysis, increase n_sims (e.g., 500)
power_results <- doas_power(
  n_sims = 10,
  surv_areas = c(2, 5), # Test scenarios: 2 DO sectors vs 5 DO sectors
  pop_size = 1000,
  mean_group = 10,
  tot_areas = 15,
  intra_group_det = 1.0,
  method = "global",
  obs_type = "independent"
)

knitr::kable(power_results, digits = 3)
```

Higher effort (more `surv_areas`) generally yields improved CI coverage and lower Coefficient of Variation (CV).

---

## 4. Understanding Base Estimators

The `doas()` function leverages several group-abundance estimators under the hood. For advanced applications, these are exported for you to use directly:

### Chapman Estimator (Independent Observers)
A bias-adjusted Lincoln-Petersen estimator.
```{r base-chapman}
chapman_estimate(both = 10, obs1_only = 2, obs2_only = 3)
chapman_variance(both = 10, obs1_only = 2, obs2_only = 3)
```

### Zippin Estimator (Dependent Observers)
A removal estimator for primary/secondary observer setups.
```{r base-zippin}
dependent_zippin_estimate(both = 10, obs1_only = 2, obs2_only = 3)
dependent_zippin_variance(both = 10, obs1_only = 2, obs2_only = 3)
```

### Bayesian Conjugate Estimator
Uses Monte Carlo simulations to draw from Beta posteriors, naturally preventing negative estimates and bounding credible intervals.
```{r base-bayesian}
set.seed(123)
bayesian_res <- bayesian_conjugate_estimate(both = 10, obs1_only = 2, obs2_only = 3, iter = 1000)
c(Mean = bayesian_res$mean, Variance = bayesian_res$var)
```
